Freezing of a Weeks-Chandler-Anderson Fluid using classical Density Functional Theory

### 2020-12-07 papers/thesis-kirstie/thesis.tex
added math to make a point explicit, edit pres

The Grand free energy is applicable to systems with a varying number of 
particles and fixed chemical potential, temperature and volume. But in 
this paper a system with a fixed temperature, volume, and number of 
particles will be of interest in which case
\begin{equation}\int n(\vec r)d\vec{r}=N\end{equation}
and so the Grand free energy is constrained
\begin{equation}\label{GrandFE}\Phi_{constrained}= F[n(\vec r)]-\mu N\end{equation}
The dependence on $n(\vec r)$ is now only present in the expression for 
the Helmholtz free energy, and so it is the Helmholtz energy as a functional 
of the number density profile $F[n(\vec r)]$ that is minimized at equilibrium. 
More specifically, in this paper, the Helmholtz free energy per particle, 
or atom,  $f_N=\frac{F}{N}$ will be minimized which is given by
\begin{equation}f_N=\frac{U}{N}-\frac{TS}{N}\end{equation}
with differential
\begin{equation}\label{usetoshowmin}df_N=d\left(\frac{U}{N}\right)-d\left(\frac{TS}{N}\right)\end{equation}
\begin{equation}df_N=-\frac{S}{N}dT-\frac{P}{N}dV-PVd\left(\frac{1}{N}\right)\end{equation}
This result shows that the natural variables of the Helmholtz free energy 
per particle $f_N$ are temperature, volume and one over the number of 
particles, which in turn also holds the number density $n=\frac{V}{N}$ fixed. 
\begin{equation}f_N(T,V, \frac{1}{N})\end{equation}

The minimization can be illustrated for a system near 
equilibrium by considering that according to the 
second law of thermodynamics, a system that is not in equilibrium will 
experience spontaneous processes that increase the total entropy of a 
system and its surroundings until equilibrium is reached, 
$\Delta{S}_{System + Surroundings} \geq 0$. Here ``surroundings" means 
the entire universe minus the system. This, together with the conservation 
of energy (the first law of thermodynamics) which says that energy lost 
(or gained) by the system is gained (or lost) by the surroundings 
$\Delta{U}_{System}=-\Delta{U}_{Surroundings}$, gives the result that 
changes in the free energy 
will be negative for spontaneous processes 
when the natural variables of the free energy are held fixed. 
For our case, with a fixed temperature, volume, and number of particles
we get
%Rewriting Eq~\ref{usetoshowmin} to show the system terms explicitly
%\begin{equation}
  %df_{System}=d\left(\frac{U}{N}\right)_{System}-Td\left(\frac{S}{N}\right)_{System}+\left(\frac{S}{N}\right)dT_{System}
%\end{equation}
%and holding the temperature fixed ($T_{System}=T_{Surroundings}$) so that 
%$dT=0$ gives
%\begin{equation}
  %f_{System}=-d\left(\frac{U}{N}\right)_{Surroundings}-Td\left(\frac{S}{N}\right)_{System}
%\end{equation}
%where conservation of energy was also applied. Fixing the volume and the 
%number of particles, so that $dV=0$ and $d\left(\frac{1}{N}\right)=0$, 
%results in
\begin{equation}
  df_{System}=-\left(\frac{T}{N}\right)dS_{System+Surroundings}
\end{equation}
%which, for fixed N, is the same as
%\begin{equation}dF_{System}=-TdS_{System+Surroundings} \end{equation} 
%Since $dS_{System+Surroundings} \geq{0}$, and the temperature and number 
%of particles are fixed, the change in the free energy of the system will 
%always be negative, and thus the free energy tends toward a minimum.
%~\cite{schroeder}. 
and so the free energy tends toward a minimum. 
Details of this procedure be found in Thermal Physics by Schroeder~\cite{schroeder}, 
and a proof of minimization in general can be found in 
Theory of Simple Fluids by Hansen and McDonald~\cite{Hansen}.

@@

Fundamental Measure Theory can be constructed from Equation~\ref{Fexcess-SPT}  
%and provides an expression for the excess Helmholtz free energy of an 
%inhomogeneous 
%hard-sphere fluid as a functional of the number density profile $n(\vec r)$. 
%
%Soft Fundamental Measure Theory (SFMT), which will be used in this paper, 
%is an extension of FMT and provides an expression for the excess Helmholtz
%free energy of fluids with soft potentials like that of the WCA fluid. 
%It will be helpful to continue looking at FMT before addressing SFMT.
%FMT begins 
by noticing that a quantity $G_{3}$ can be defined such that
\begin{equation}{\eta =\left(\frac{N}{V}\right)\frac{4}{3}\pi{R^3}=nG_{3}}\end{equation}
where $G_{3}$ is the volume of a sphere. 
Likewise, by defining quantities
\begin{equation}{G_{2}=4\pi{R^2}}\end{equation}
\begin{equation}{G_{1}=R}\end{equation}
\begin{equation}{G_{0}=1}\end{equation}
where $G_{2}$ is the surface area of a sphere, and $G_{1}$ is the 
radius of a sphere, Equation~\ref{Fexcess-SPT} can be written as
\begin{equation}\label{FexfromSPTinthermsofG}{F_{excess}=k_{B}T\left(-nG_{0}ln(1-nG_{3})+\frac{nG_{1}nG_{2}}{1-nG_{3}}+\frac{nG_{2}^3}{24\pi(1-Gn_{3})^2}\right)V}\end{equation}
The quantities $G_{3}, G_{2}, G_{1},$ and $G_{0}$ are fundamental 
geometric measures of a sphere, and are what give rise to the name
Fundamental Measure Theory. Introducing a set of variables
%weighted densities $n_{i}=nG_{i}$ 
given by
\begin{equation}\label{n3}{n_{3}=nG_{3}=\left(\frac{N}{V}\right)\frac{4}{3}\pi{R^3}=\eta}\end{equation}
\begin{equation}\label{n2}{n_{2}=nG_{2}=\left(\frac{N}{V}\right)4\pi{R^2}=\frac{3\eta}{R}}\end{equation}
\begin{equation}\label{n1}{n_{1}=nG_{1}=\left(\frac{N}{V}\right)R}\end{equation}
\begin{equation}\label{n0}{n_{0}=nG_{0}=\left(\frac{N}{V}\right)}\end{equation}
Equation~\ref{FexfromSPTinthermsofG} can be written as
\begin{align}\label{FexfromSPT}
    F_{excess}=k_{B}T\left(-n_{0}ln(1-n_{3})+\frac{n_{1}n_{2}}{1-n_{3}}+\frac{n_{2}^3}{24\pi(1-n_{3})^2}\right)V
\end{align}
which has the form 
\begin{align}
    F_{excess} = k_BT(\Phi_1+\Phi_2+\Phi_3)V
\end{align}
where 
\begin{align}
   \Phi_1 &= -n_{0}ln(1-n_{3}) \\
   \Phi_2 &= \frac{n_{1}n_{2}}{1-n_{3}} \\
   \Phi_3 &= \frac{n_{2}^3}{24\pi(1-n_{3})^2}   \label{oldPhi3equation}
\end{align}

>>>>

The Grand free energy is applicable to systems with a varying number of 
particles and fixed chemical potential, temperature and volume. But in 
this paper a system with a fixed temperature, volume, and number of 
particles will be of interest in which case
\begin{equation}\int n(\vec r)d\vec{r}=N\end{equation}
and so the Grand free energy is constrained
\begin{equation}\label{GrandFE}\Phi_{constrained}= F[n(\vec r)]-\mu N\end{equation}
The dependence on $n(\vec r)$ is now only present in the expression for 
the Helmholtz free energy, and so it is the Helmholtz energy as a functional 
of the number density profile $F[n(\vec r)]$ that is minimized at equilibrium. 
More specifically, in this paper, the Helmholtz free energy per particle, 
or atom,  $f_N=\frac{F}{N}$ will be minimized which is given by
\begin{equation}f_N=\frac{U}{N}-\frac{TS}{N}\end{equation}
with differential
\begin{equation}\label{usetoshowmin}df_N=d\left(\frac{U}{N}\right)-d\left(\frac{TS}{N}\right)\end{equation}
\begin{equation}df_N=-\frac{S}{N}dT-\frac{P}{N}dV-PVd\left(\frac{1}{N}\right)\end{equation}
This result shows that the natural variables of the Helmholtz free energy 
per particle $f_N$ are temperature, volume and one over the number of 
particles, which in turn also holds the number density $n=\frac{V}{N}$ fixed. 
\begin{equation}f_N(T,V, \frac{1}{N})\end{equation}
Taking the differential
\begin{displaymath}df_N=\frac{\partial f_N}{\partial T}dT +\frac{\partial f_N}{V}dV+ \frac{\partial f_N}{\partial \left(\frac{1}{N}\right)}d\left(\frac{1}{N}\right)\end{displaymath}
and using the relation
\begin{displaymath}d\left(\frac{1}{n}\right)=Vd\left(\frac{1}{N}\right)+\frac{1}{N}dV\end{displaymath}
we get
\begin{displaymath}df_N=\frac{\partial f_N}{\partial T}dT + \frac{\partial f_N}{\partial \frac{1}{n}}d\left(\frac{1}{n}\right)\end{displaymath}

\begin{equation}f_N(T,\frac{1}{n})\end{equation}

The minimization can be illustrated for a system near 
equilibrium by considering that according to the 
second law of thermodynamics, a system that is not in equilibrium will 
experience spontaneous processes that increase the total entropy of a 
system and its surroundings until equilibrium is reached, 
$\Delta{S}_{System + Surroundings} \geq 0$. Here ``surroundings" means 
the entire universe minus the system. This, together with the conservation 
of energy (the first law of thermodynamics) which says that energy lost 
(or gained) by the system is gained (or lost) by the surroundings 
$\Delta{U}_{System}=-\Delta{U}_{Surroundings}$, gives the result that 
changes in the free energy 
will be negative for spontaneous processes 
when the natural variables of the free energy are held fixed.

For our case, with a fixed temperature and number density we get
%Rewriting Eq~\ref{usetoshowmin} to show the system terms explicitly
%\begin{equation}
  %df_{System}=d\left(\frac{U}{N}\right)_{System}-Td\left(\frac{S}{N}\right)_{System}+\left(\frac{S}{N}\right)dT_{System}
%\end{equation}
%and holding the temperature fixed ($T_{System}=T_{Surroundings}$) so that 
%$dT=0$ gives
%\begin{equation}
  %f_{System}=-d\left(\frac{U}{N}\right)_{Surroundings}-Td\left(\frac{S}{N}\right)_{System}
%\end{equation}
%where conservation of energy was also applied. Fixing the volume and the 
%number of particles, so that $dV=0$ and $d\left(\frac{1}{N}\right)=0$, 
%results in
\begin{equation}
  df_{System}=-\left(\frac{T}{N}\right)dS_{System+Surroundings}
\end{equation}
%which, for fixed N, is the same as
%\begin{equation}dF_{System}=-TdS_{System+Surroundings} \end{equation} 
%Since $dS_{System+Surroundings} \geq{0}$, and the temperature and number 
%of particles are fixed, the change in the free energy of the system will 
%always be negative, and thus the free energy tends toward a minimum.
%~\cite{schroeder}. 
and so the free energy tends toward a minimum. 
Details of this procedure be found in Thermal Physics by Schroeder~\cite{schroeder}, 
and a proof of minimization in general can be found in 
Theory of Simple Fluids by Hansen and McDonald~\cite{Hansen}.

@@

Fundamental Measure Theory can be constructed from Equation~\ref{Fexcess-SPT}  
%and provides an expression for the excess Helmholtz free energy of an 
%inhomogeneous 
%hard-sphere fluid as a functional of the number density profile $n(\vec r)$. 
%
%Soft Fundamental Measure Theory (SFMT), which will be used in this paper, 
%is an extension of FMT and provides an expression for the excess Helmholtz
%free energy of fluids with soft potentials like that of the WCA fluid. 
%It will be helpful to continue looking at FMT before addressing SFMT.
%FMT begins 
by noticing that a quantity $G_{3}$ can be defined such that
\begin{equation}{\eta =\left(\frac{N}{V}\right)\frac{4}{3}\pi{R^3}=nG_{3}}\end{equation}
where $G_{3}$ is the volume of a sphere. 
Likewise, by defining quantities
\begin{equation}{G_{2}=4\pi{R^2}}\end{equation}
\begin{equation}{G_{1}=R}\end{equation}
\begin{equation}{G_{0}=1}\end{equation}
where $G_{2}$ is the surface area of a sphere, and $G_{1}$ is the 
radius of a sphere, Equation~\ref{Fexcess-SPT} can be written as
\begin{equation}\label{FexfromSPTinthermsofG}{F_{excess}=k_{B}T\left(-nG_{0}ln(1-nG_{3})+\frac{nG_{1}nG_{2}}{1-nG_{3}}+\frac{nG_{2}^3}{24\pi(1-Gn_{3})^2}\right)V}\end{equation}
The quantities $G_{3}, G_{2}, G_{1},$ and $G_{0}$ are fundamental 
geometric measures of a sphere, and are what give rise to the name
Fundamental Measure Theory. Introducing a set of variables
%weighted densities $n_{i}=nG_{i}$ 
given by
\begin{equation}\label{n3}{n_{3}=nG_{3}=\left(\frac{N}{V}\right)\frac{4}{3}\pi{R^3}=\eta}\end{equation}
\begin{equation}\label{n2}{n_{2}=nG_{2}=\left(\frac{N}{V}\right)4\pi{R^2}=\frac{3\eta}{R}}\end{equation}
\begin{equation}\label{n1}{n_{1}=nG_{1}=\left(\frac{N}{V}\right)R}\end{equation}
\begin{equation}\label{n0}{n_{0}=nG_{0}=\left(\frac{N}{V}\right)}\end{equation}
Equation~\ref{FexfromSPTinthermsofG} can be written as
\begin{align}\label{FexfromSPT}
    F_{excess}=k_{B}T\left(-n_{0}ln(1-n_{3})+\frac{n_{1}n_{2}}{1-n_{3}}+\frac{n_{2}^3}{24\pi(1-n_{3})^2}\right)V
\end{align}
which has the form 
\begin{align}
    F_{excess} = k_BT(\Phi_1+\Phi_2+\Phi_3)V
\end{align}
%where 
\begin{align}
   \Phi_1 &= -n_{0}ln(1-n_{3}) \\
   \Phi_2 &= \frac{n_{1}n_{2}}{1-n_{3}} \\
   \Phi_3 &= \frac{n_{2}^3}{24\pi(1-n_{3})^2}   \label{oldPhi3equation}
\end{align}
